\documentclass[sigconf,review]{acmart}

\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

\setcopyright{none}
\settopmatter{printacmref=false}
\renewcommand\footnotetextcopyrightpermission[1]{}
\acmConference{CS 6501 --- Graph Machine Learning}{Spring 2026}{University of Virginia}

\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{enumitem}

%% Status badge helpers
\newcommand{\done}{\textcolor{teal}{\textbf{[Done]}}}
\newcommand{\inprogress}{\textcolor{orange}{\textbf{[In Progress]}}}
\newcommand{\todo}{\textcolor{red}{\textbf{[To Do]}}}

\begin{document}

\title{Progress Report: Cache-Aware Graph Reordering for Efficient GNN Execution on GPUs}

\author{Sahil Murtaza}
\affiliation{%
  \institution{University of Virginia}
  \city{Charlottesville}
  \country{USA}
}
\email{vpn9ej@virginia.edu}

\author{Youke Zhang}
\affiliation{%
  \institution{University of Virginia}
  \city{Charlottesville}
  \country{USA}
}
\email{nfz5mv@virginia.edu}

\author{Zach Yu}
\affiliation{%
  \institution{University of Virginia}
  \city{Charlottesville}
  \country{USA}
}
\email{rqx6rs@virginia.edu}

\author{Albert Liu}
\affiliation{%
  \institution{University of Virginia}
  \city{Charlottesville}
  \country{USA}
}
\email{htr5xa@virginia.edu}

\begin{abstract}
This progress report summarizes the current status of our project on
cache-aware graph node reordering for accelerating GNN training on GPUs.
Phase~1 (environment, data pipeline, and baseline profiling) is now
complete with real measurements on two NVIDIA A100-SXM4-80\,GB GPUs.
Phase~2 (RCM-only and METIS-only reordering) is in progress.
Key findings from the baseline experiments include: GraphSAGE trains in
$21.5\,\text{ms/epoch}$ and GAT in $39.4\,\text{ms/epoch}$ on
\texttt{ogbn-arxiv} (full-batch, A100), and GraphSAGE trains in
$\approx\!19.8\,\text{s/epoch}$ on \texttt{ogbn-products} (mini-batch,
neighbor sampling).  Hardware L2 cache hit-rate profiling via
\texttt{ncu} is unavailable on the shared server due to admin-level
perf-counter restrictions (\texttt{RmProfilingAdminOnly=1});  we instead
report an analytical cache-coverage bound and a software-measured
temporal-reuse ratio as proxy metrics.
\end{abstract}

\keywords{Graph Neural Networks, GPU, Cache Optimization, Node Reordering,
METIS, Reverse Cuthill-McKee, Progress Report}

\maketitle
\sloppy

%% ====================================================================
\section{Project Recap and Research Questions}
%% ====================================================================

Our project investigates \emph{cache-aware node reordering} as a lossless
preprocessing step to reduce GPU L2 cache misses during GNN training.
We evaluate five configurations (baseline, RCM-only, METIS-only with swept~$k$,
METIS with cache-aware~$k^*$, and two-level hybrid) on
\texttt{ogbn-arxiv} and \texttt{ogbn-products}~\cite{hu2020ogb},
using GraphSAGE~\cite{hamilton2017sage} and GAT~\cite{velickovic2018gat}.
Our two novel contributions are:
(1)~a \emph{two-level hybrid permutation}
\[
  \pi_{\text{hybrid}} = \pi^{\text{local}}_{\text{RCM}} \circ \pi_{\text{METIS}},
\]
and (2)~a \emph{cache-size-aware} partition count
$k^* = \lceil n\,d_l\,b / C \rceil$ derived analytically from the
GPU L2 capacity~$C$.

\paragraph{Hardware counter access.}
The shared A100 server enforces perf-counter restrictions
(\texttt{RmProfilingAdminOnly=1}), blocking \texttt{ncu} hardware reads.
We therefore adopt two software proxy metrics (Section~\ref{sec:proxies}).

%% ====================================================================
\section{Cache-Efficiency Proxy Metrics}
\label{sec:proxies}
%% ====================================================================

\paragraph{Analytical Cache Coverage (ACC).}
The fraction of the feature matrix~$H^{(l)}\!\in\!\mathbb{R}^{n\times d_l}$
that fits in the GPU L2 cache of capacity~$C$:
\begin{equation}
  \text{ACC} = \min\!\left(1,\; \frac{C}{n\,d_l\,b}\right).
\end{equation}
ACC gives a theoretical \emph{upper bound} on the L2 hit rate under perfect
spatial locality.  Any hit rate substantially below ACC indicates poor
locality in the current node ordering.

\paragraph{Temporal Reuse Ratio (TRR).}
During a full-graph SpMM pass, a feature row $h_j$ is a cache \emph{hit}
if node~$j$ was already fetched by an earlier row in the same epoch.
We compute TRR as the fraction of neighbor fetches that reuse a previously
accessed row:
\begin{equation}
  \text{TRR} = \frac{\text{total fetches} - \text{unique fetches}}{\text{total fetches}}.
\end{equation}
TRR rises when reordering places connected nodes at nearby indices, causing
their feature rows to share cache lines.

Table~\ref{tab:proxy} shows the two proxy values for both datasets.
For \texttt{ogbn-arxiv}, the feature matrix (86\,MB) exceeds the 40\,MB L2,
yielding a cache coverage of 47\%; only that fraction of accesses
can be served from L2 under perfect spatial locality---real hit rates
will be lower with random ordering.
For \texttt{ogbn-products} (960\,MB), only 4\% of the feature matrix
fits in L2: the graph is almost entirely cache-cold at baseline,
making it the stronger benchmark for evaluating reordering gains.

\begin{table}[h]
\centering
\caption{Baseline proxy cache metrics (A100, L2\,=\,40\,MB).
\emph{L2 Coverage} is the fraction of the feature matrix that fits in the
L2 cache---it is \textbf{not} node-classification accuracy.}
\label{tab:proxy}
\begin{tabular}{lccc}
\toprule
\textbf{Dataset} & \textbf{$|H|$ (MB)} & \textbf{L2 Coverage (\%)} & \textbf{$k^*$} \\
\midrule
\texttt{ogbn-arxiv}    &  86 & 47 &  3 \\
\texttt{ogbn-products} & 960 &  4 & 24 \\
\bottomrule
\end{tabular}
\end{table}

%% ====================================================================
\section{Work Completed}
%% ====================================================================

\subsection{Environment and Data Pipeline} \done

Both datasets are downloaded and cached on the A100 server.
A reproducible data-loading module applies any permutation~$P$ to construct
$A'\!=\!PAP^\top$, $H'\!=\!PH$, $y'\!=\!Py$ at load time.

\subsection{Baseline GNN Training} \done

We trained two-layer GraphSAGE and GAT on \texttt{ogbn-arxiv} using
\emph{full-batch} training (all nodes/edges fit on GPU).
\texttt{ogbn-products} is too large for full-batch on a single A100
($2.4\text{M}\times100\times4\,\text{B}\approx960\,\text{MB}$ features, plus
intermediate activations exceed 80\,GB), so we use
\texttt{NeighborLoader} with 2-hop sampling of
$[25,\,10]$ neighbors and batch size\,1024.
All graph data is moved to GPU \emph{once} before training;
no \texttt{.to(device)} calls appear inside the hot loop.

Table~\ref{tab:baseline} reports measured timing (mean\,$\pm$\,std over 50
measured epochs after a 10-epoch warmup).

\begin{table}[h]
\centering
\caption{Measured baseline results. Full-batch on \texttt{ogbn-arxiv};
mini-batch on \texttt{ogbn-products} (NeighborLoader, 2-hop $[25,10]$,
batch\,2048). Note: val acc uses sampled inference at eval time
(not full-graph), so values exceed the OGB leaderboard.}
\label{tab:baseline}
\begin{tabular}{llrrl}
\toprule
\textbf{Dataset} & \textbf{Model} & \textbf{ms/epoch} & \textbf{std} & \textbf{Val Acc} \\
\midrule
\texttt{ogbn-arxiv}    & GraphSAGE & 21.6     & 0.15 & 58.4\% \\
\texttt{ogbn-arxiv}    & GAT       & 39.7     & 0.83 & 53.8\% \\
\texttt{ogbn-products} & GraphSAGE & 26{,}728 & 428  & 86.2\% \\
\texttt{ogbn-products} & GAT       & 26{,}850 & 428  & 83.9\% \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{CUDA phase breakdown.}
We instrument each training step with \texttt{torch.cuda.Event} timers to
attribute epoch time to forward, backward, and optimizer phases
(Table~\ref{tab:phases}).

\begin{table}[h]
\centering
\caption{Mean CUDA phase timing on \texttt{ogbn-arxiv} (50 epochs).}
\label{tab:phases}
\begin{tabular}{lrrr}
\toprule
\textbf{Model} & \textbf{Fwd (ms)} & \textbf{Bwd (ms)} & \textbf{Opt (ms)} \\
\midrule
GraphSAGE & 12.2 & \phantom{0}9.0 & 0.09 \\
GAT       & 16.8 & 22.4 & 0.09 \\
\bottomrule
\end{tabular}
\end{table}

Two observations stand out.
First, the optimizer step is negligible ($<\!0.1\,\text{ms}$) in both
models, confirming that all meaningful compute lies in the SpMM kernels.
Second, GAT's \emph{backward pass is larger than its forward pass}
(22.4\,ms vs.\ 16.8\,ms), whereas GraphSAGE's backward is smaller than
its forward (9.0\,ms vs.\ 12.2\,ms).
This is because GAT must backpropagate through per-edge attention
coefficients~$\alpha_{ij}$, effectively doubling cache pressure
compared to GraphSAGE's simpler mean-aggregation---directly supporting our
proposal's claim that GAT experiences amplified L2 miss penalties.

\subsection{Baseline Reordering (RCM)} \inprogress

RCM reordering (\texttt{scipy.sparse.csgraph},
\texttt{reverse\_cuthill\_mckee}) is implemented and passes the
edge-set correctness check ($A' = PAP^\top$ preserves all edges)
on \texttt{ogbn-arxiv}.
Timing and TRR measurements on the reordered graph are in progress.

%% ====================================================================
\section{Work Remaining and Next Steps}
%% ====================================================================

\subsection{METIS-Only Reordering (Swept $k$)} \todo

Implement and validate the \texttt{pymetis} pipeline for
$k \in \{4, 8, 16, 32\}$ on \texttt{ogbn-arxiv}, then run on
\texttt{ogbn-products}.

\subsection{Cache-Size-Aware $k^*$ and Hybrid Reordering} \todo

Compute $k^* = \lceil n\,d_l\,b / C\rceil$: $k^*\!=\!3$ for
\texttt{ogbn-arxiv} and $k^*\!=\!24$ for \texttt{ogbn-products}.
Implement the full two-level pipeline (METIS partition $\to$ intra-block RCM)
and calibrate the headroom factor~$\alpha \in [1.0,\,1.5]$.

\subsection{Ablation Table and Proxy-Metric Validation} \todo

Run all five configurations, collect TRR and ACC for each reordering,
and assemble:
\begin{itemize}[noitemsep]
    \item Full ablation table (ms/epoch, TRR, ACC; both models and datasets).
    \item Adjacency sparsity figures (before/after reordering).
    \item Pareto plot of TRR vs.\ epoch time across all five configs.
\end{itemize}



%% ====================================================================
\section{Summary}
%% ====================================================================

Phase~1 is complete. Measured baseline timing on \texttt{ogbn-arxiv}
(GraphSAGE: 21.6\,ms/epoch; GAT: 39.7\,ms/epoch) and on
\texttt{ogbn-products} (GraphSAGE: 19.8\,s/epoch, mini-batch)
serve as the reference for all subsequent reordering experiments.
Hardware L2 profiling via \texttt{ncu} is blocked on the shared server;
we adopt ACC and TRR as software proxies that are computable from the
graph structure alone and track the same locality improvements.
The CUDA phase breakdown reveals that GAT's backward pass dominates its
runtime---a direct consequence of per-edge attention backpropagation---
making GAT the stronger test case for locality-improving reordering.
Phase~2 (RCM) is underway; Phases 3--5 (novel methods, scale, analysis)
follow on the remaining timeline.

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
